# ---
title: "Simulating realistic microbial observations with SparseDOSSA2"
author:
- name: "Siyuan Ma"
affiliation:
- Harvard T.H. Chan School of Public Health
- Broad Institute
email: syma.research@gmail.com
package: SparseDOSSA2
date: "12/01/2020"
output:
BiocStyle::html_document
vignette: >
%\VignetteIndexEntry{SparseDOSSA2}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
bibliography: references.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```

# Introduction

`SparseDOSSA2` an R package for fitting to and the simulation of realistic microbial
abundance observations. It provides functionlaities for: a) generation of realistic synthetic microbial observations, b) spiking-in of associations with metadata variables 
for e.g. benchmarking or power analysis purposes, and c) fitting the SparseDOSSA 2 
model to real-world microbial abundance observations that can be used for a). This vignette is intended to provide working examples for these functionalities. 

```{r, message=FALSE}
library(SparseDOSSA2)
# tidyverse packages for utilities
library(magrittr)
library(dplyr)
library(ggplot2)
```

# Installation

SparseDOSSA2 is a Bioconductor package and can be installed via the following 
command.

```{r Installation, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("SparseDOSSA2")
```

# Simulating realistic microbial observations with `SparseDOSSA2`
The most important functionality of `SparseDOSSA2` is the simulation of
realistic synthetic microbial observations. To this end, `SparseDOSSA2` provides
three pre-trained templates, `"Stool"`, `"Vaginal"`, and `"IBD"`, targeting
continuous, discrete, and diseased population structures. 
```{r SparseDOSSA2_generate}
Stool_simulation <- SparseDOSSA2(template = "Stool", 
                                 n_sample = 100, 
                                 n_feature = 100,
                                 verbose = TRUE)
Vaginal_simulation <- SparseDOSSA2(template = "Vaginal", 
                                   n_sample = 100, 
                                   n_feature = 100,
                                   verbose = TRUE)
```

# Fitting to microbiome datasets with SparseDOSSA2

SparseDOSSA2 provide two functions, `fit_SparseDOSSA2` and `fitCV_SparseDOSSA2`,
to fit the SparseDOSSA2 model to microbial count or relative abundance observations.
For these functions, as input, `SparseDOSSA2` requires a feature-by-sample 
table of microbial abundance observations. We provide with SparseDOSSA2 a minimal
example of such a dataset: a five-by-five of the [HMP1-II](https://www.hmpdacc.org/hmp/) 
stool study.

```{r load data}
data("Stool_subset", package = "SparseDOSSA2")
# columns are samples.
Stool_subset[1:2, 1, drop = FALSE]
```

## Fitting SparseDOSSA2 model with `fit_SparseDOSSA2`

`fit_SparseDOSSA2` fits the SparseDOSSA2 model to estimate the 
model parameters: per-feature prevalence, mean and standard deviation of 
non-zero abundances, and feature-feature correlations. 
It also estimates joint distribution of these parameters
and (if input is count) a read count distribution.

```{r SparseDOSSA2_fit}
fitted <- fit_SparseDOSSA2(data = Stool_subset,
                           control = list(verbose = TRUE))
# fitted mean log non-zero abundance values of the first two features
fitted$EM_fit$fit$mu[1:2]
```

## Fitting SparseDOSSA2 model with `fitCV_SparseDOSSA2`

The user can additionally achieve optimal model fitting via 
`fitCV_SparseDOSSA2`. They can either provide a vector of tuning parameter
values (`lambdas`) to control sparsity in the estimation of the correlation
matrix parameter, or a grid will be selected automatically. 
`fitCV_SparseDOSSA2` uses cross validation to select an "optimal" model fit
across these tuning parameters via average testing log-likelihood. This is a 
computationally intensive procedure, and best-suited for users that would like
accurate fitting to the input dataset, for best simulated new microbial 
observations on the same features as the input (i.e. not new features).

 ```{r SparseDOSSA2_fit_CV}
# set.seed(1)
# fitted_CV <- fitCV_SparseDOSSA2(data = Stool_subset,
#                                         lambdas = c(0.1, 1),
#                                         K = 2,
#                                         control = list(verbose = TRUE))
# the average log likelihood of different tuning parameters
# apply(fitted_CV$EM_fit$logLik_CV, 2, mean)
# The second lambda (1) had better performance in terms of log likelihood,
# and will be selected as the default fit.
# ```

## Parallelization controls with `future`

#`SparseDOSSA2` internally uses `r BiocStyle::CRANpkg("future")` to allow for
#parallel computation. The user can thus specify parallelization through `future`'s
#interface. See the [reference #manual](https://cran.r-project.org/web/packages/future/vignettes/future-1-overview.html) 
#for `future` for more details. This is 
#particularly suited if fitting `SparseDOSSA2` in a high-performance computing
#environment/

#```{r future}
## regular fitting 
# system.time(fitted_regular <- 
#               fit_SparseDOSSA2(data = Stool_subset,
#                                control = list(verbose = FALSE)))
## parallel fitting with future:
# future::plan(future::multisession())
# system.time(fitted_parallel <- 
#               fit_SparseDOSSA2(data = Stool_subset,
#                                control = list(verbose = FALSE)))

## For CV fitting, there are three components that can be paralleled, in order:
## different cross validation folds, different tuning parameter lambdas, 
## and different samples. It is usually most efficient to parallelize at the
## sample level:
# system.time(fitted_regular_CV <-
#               fitCV_SparseDOSSA2(data = Stool_subset,
#                                  lambdas = c(0.1, 1),
#                                  K = 2,
#                                  control = list(verbose = TRUE)))
# future::plan(future::sequential(), future::sequential(), future::multisession())
# system.time(fitted_parallel_CV <-
#               fitCV_SparseDOSSA2(data = Stool_subset,
#                                  lambdas = c(0.1, 1),
#                                  K = 2,
#                                  control = list(verbose = TRUE)))

```


# Sessioninfo
```{r sessioninfo}
sessionInfo()
```
# References