<!-- # --- -->
<!-- title: "Simulating realistic microbial observations with SparseDOSSA2" -->
<!-- author:  -->
<!-- - name: "Siyuan Ma" -->
  <!-- affiliation:  -->
  <!-- - Harvard T.H. Chan School of Public Health -->
  <!-- - Broad Institute -->
  <!-- email: siyuanma@g.harvard.edu -->
<!-- package: SparseDOSSA2 -->
<!-- date: "07/10/2020" -->
<!-- output:  -->
  <!-- BiocStyle::html_document -->
<!-- vignette: > -->
  <!-- %\VignetteIndexEntry{SparseDOSSA2} -->
  <!-- %\VignetteEngine{knitr::rmarkdown} -->
  <!-- %\VignetteEncoding{UTF-8}   -->
<!-- bibliography: references.bib -->
<!-- --- -->

<!-- # ```{r setup, include=FALSE} -->
<!-- knitr::opts_chunk$set(echo = TRUE) -->
<!-- knitr::opts_chunk$set(cache = FALSE) -->
<!-- ``` -->

# Introduction

`SparseDOSSA2` is an R package for fitting and simulation of realistic microbial abundance observations.

```{r, message=FALSE}
library(SparseDOSSA2)
# tidyverse packages for utilities
library(magrittr)
library(dplyr)
library(ggplot2)
```

# Installation

SparseDOSSA2 is available through GitHub as of now and can be installed via the 
following command.

```{r Installation, eval = FALSE}
devtools::install_github("biobakery/SparseDOSSA2")
```

# Input data

As input, `SparseDOSSA2` requires a feature-by-sample abundance table of microbial 
abundance observations. They can be either count or on the relative abundance
scale.

```{r load data}
data("HMP_stool", package = "SparseDOSSA2")
# columns are samples.
stool[1:5, 1, drop = FALSE]
```

# Fitting SparseDOSSA2 model with `fit_SparseDOSSA2`

`fit_SparseDOSSA2` fits the SparseDOSSA2 hierarchical model to estimate the 
model parameters: per-feature prevalence, and mean and standard deviation of 
non-zero abundances. It also estimates joint distribution of these parameters
and (if input is count) a read count distribution.

```{r SparseDOSSA2_fit}
SparseDOSSA2_fit <- fit_SparseDOSSA2(data = stool,
                                     control = list(verbose = TRUE))
names(SparseDOSSA2_fit)
```

# Fitting SparseDOSSA2 model with `fitCV_SparseDOSSA2`

The user can additionally achieve optimal model fitting via 
`fitCV_SparseDOSSA2`. They can either provide a vector of tuning parameter
values (`lambdas`) to control sparsity in the estimation of the correlation
matrix parameter, or a grid will be selected automatically. 
`fitCV_SparseDOSSA2` uses cross validation to select an "optimal" model fit
across these tuning parameters via average testing log-likelihood. This is a 
computationally intensive procedure, and best-suited for users that would like
accurate fitting to the input dataset, for best simulated new microbial 
observations on the same features as the input (i.e. not new features).

```{r SparseDOSSA2_fit_CV}
set.seed(1)
SparseDOSSA2_CVfit <- fitCV_SparseDOSSA2(data = stool,
                                         lambdas = c(0.1, 1),
                                         K = 2,
                                         control = list(verbose = TRUE))
names(SparseDOSSA2_CVfit)
```

# Simulating realistic microbial observations with `SparseDOSSA2`
```{r SparseDOSSA2_generate}
SparseDOSSA2_simulation <- SparseDOSSA2(template = SparseDOSSA2_fit,
                                        new_features = TRUE,
                                        n_sample = 100,
                                        n_feature = 200,
                                        spike_metadata = "abundance",
                                        control = list(verbose = TRUE))
names(SparseDOSSA2_simulation)
head(SparseDOSSA2_simulation$spike_metadata$feature_metadata_spike_df)
```

# Parallelization controls with `future`

```{r future}
# future::plan(future::sequential, future::multicore)
# fit_SparseDOSSA2 <- SparseDOSSA2_fit(data = stool,
#                                      control = list(verbose = TRUE))
# 
# future::plan(future::sequential, future::sequential, future::multicore)
# fit_SparseDOSSA2 <- SparseDOSSA2_fit_CV(data = stool,
#                                         lambdas = c(0.1, 1),
#                                         K = 2,
#                                         control = list(verbose = TRUE))
```

# Sessioninfo
```{r sessioninfo}
sessionInfo()
```
# References