---
title: "Numeric integrations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/n/janson_lab/lab/sma/sparsedossa_update/")
knitr::opts_chunk$set(echo=FALSE)
library(magrittr)
library(ggplot2)
```

```{r load data, cache=TRUE, include=FALSE}
smar::sourceDir("SparseDOSSA2/R/")
dir_output <- "results/integrations/"
load(paste0(dir_output, "tb_job.RData"))
tb_job <- tb_job %>% 
  dplyr::group_by(i_job) %>% 
  dplyr::mutate(l_integrations =
                  {
                    load(paste0(dir_output, "fit_", i_job, ".RData"))
                    list(fit_integrate)
                  })
tb_job <- tb_job %>% 
  dplyr::mutate(integral =
                  l_integrations %>% 
                  purrr::map_dbl("integral"),
                abs_error =
                  l_integrations %>% 
                  purrr::map_dbl("error"),
                n_eval =
                  l_integrations %>% 
                  purrr::map_dbl("neval")) %>% 
  dplyr::mutate(rel_error = abs_error / integral)
```

$f_{\log A_s, X}$ as a function of $a_s$ takes roughly a bell shape. Two issues:

1. How to integrate (i.e. where to evaluate the function)? 

2. how to choose integration limits?

``` {r normal curve}
plot(seq(-5, 5, by = 0.2), dnorm(seq(-5, 5, by = 0.2)))
```

Three integration approaches:

* Adaptive Gaussian quadrature, implemented in 
  [`cubature`](https://github.com/stevengj/cubature)
    
    * 15 fixed knots, exact evaluation of polynomials up to 29 degrees
    
    * Adaptively sub-divide regions to improve estimation
    
    * Knots are not re-used
    
* Clenshaw–Curtis quadrature, also implemented in 
  [`cubature`](https://github.com/stevengj/cubature)
  
    * Adaptively doubles the number of knots to improve estimation
    
    * Knots are re-used
  
* Use cubic spline for approximating the curve + integration

    * Errors for each segment can be estimated by monotonicity
  
    * Adaptively add knots to regions with largest error estimation to improve
    estimation
    
    * Knots are re-used
  
For integration limits, fit a rough normal density to the curve based on
small number of points (~10), then use quantiles (1% and 99%, for example) as
limits!

```{r plotting, fig.width=15, fig.height=10}
tb_job %>% 
  dplyr::filter(i_param <= 5) %>% 
  ggplot(aes(x = log2(n_eval), y = integral, color = method,
             shape = as.factor(limit_tol))) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ i_sample + i_param, scales = "free_y",
             ncol = 5) +
  ggtitle("Integrals")

tb_job %>% 
  dplyr::filter(i_param <= 5) %>% 
  ggplot(aes(x = log2(n_eval), y = log10(rel_error), color = method,
             shape = as.factor(limit_tol))) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~ i_sample + i_param, scales = "free_y",
             ncol = 5)
```