---
title: "Likelihoods"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
library(magrittr)
library(ggplot2)
```

```{r load data, include=FALSE}
load("../data/physeqs/genera.RData")
physeq <- physeq_genera %>% 
  phyloseq::subset_samples(dataset_name == "MucosalIBD") %>% 
  smar::prune_taxaSamples(flist_taxa = genefilter::kOverA(k = 5, A = 1))
mat_X_count <- t(smar::otu_table2(physeq))
colnames(mat_X_count) <- CRTmicrobiome:::simplify_feature(colnames(mat_X_count))
mat_X <- t(apply(mat_X_count, 1, function(x) x / sum(x)))
```

```{r whats wrong with likelihood}
load("../results/EM_diagnostics/mcmc.RData")

tmp <- sapply(seq_len(nrow(mat_X)),
              function(i) {
                dx(x = mat_X[i, ],
                   pi0 = fit_EM_mcmc$ll_params[[20]]$pi0,
                   mu = fit_EM_mcmc$ll_params[[20]]$mu,
                   sigma = fit_EM_mcmc$ll_params[[20]]$sigma,
                   Omega = fit_EM_mcmc$ll_params[[20]]$Omega)$abs.error
              })
dmultinom(mat_X_count[1, ], prob = mat_X[1, ])
samples <- rtest(n = 10000, x = mat_X_count[1, ])

probs <- vapply(seq_len(nrow(samples)),
                function(i) dmultinom(mat_X_count[1, ], prob = samples[i, ]),
                0.0)
rtest <- function(n, x) {
  size <- sum(x)
  x_pseudo <- ifelse(x > 0,
                     x / size,
                     (x + 0.5) / size)
  mean_norm <- log(x_pseudo)
  sd_norm <- sqrt((1 - x_pseudo) / x_pseudo / size)
  samples <- vapply(seq_along(x), 
                    function(j) {
                      rj <- exp(rnorm(n = n, 
                                      mean = mean_norm[j],
                                      sd = sd_norm[j]))
                      if(x[j] == 0)
                        rj <- rj * rbinom(n = n, size = 1, prob = 0.5)
                      return(rj)
                    },
                    rep(0.0, n))
  return(samples / apply(samples, 1, sum))
}

```

## Numeric calculation of likelihood
* Goal: calculation $f_C(c | \theta)$

\begin{align}
f_C(c | \theta) = \int f_C(c | x) f_X(x | \theta) dx
\end{align}

* Problems and solutions:
  * $f_C(c | x)$ multinomial, and as function of $x$ is concentrated around 
    $\hat{x} = \frac{c}{\sum c}$
    
    * Solution: propose $x$ from densitiy $g(x)$ around $\hat{x}$. Estimate $f_C(c | \theta)$ as
      $E_g \frac{f_C(c | x)}{g(x)} f_X(x | \theta)$
  * $g$ has restrictions:
    
    * Must be zero-inflated ($0 < P_g(x_j = 0) < 1$) since $f_X$ is 
      zero-inflated
    * Must be compositional since $f_X$ is compositional
    * Solution: propose $\tilde{x}_j$ from normal $\times$ bernoulli, then let
      $x = \frac{\tilde{x}}{\sum{\tilde{x}}}$
  * Computation
    
    * For each $c$, need $R$ Monte Carlo samples of $x$ from $g$ to calculate 
      $E_g$
    * For each $x$, need $R$ Monte Carlo samples to calculate $g(x)$ and 
      $f_X(x | \theta)$
    * Computation cost is $n * R^2$